---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Buchanan, Samuel"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks,"" and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded.  Points assigned to each item appear in this template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections ##########################

##### Section 1: (6 points) Summarizing the data.

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
table(mydata$CLASS, mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (The sex category has roughly equal numbers of observations, which is good for potential sampling studies later on. Likewise the different classes seem to have a roughly normal distribution of counts, which is also good for later analyses. There also seem to be a few outliers in the RINGS and SHUCK categories, since the max value is several times more than the 3rd quantile. But for the most part, the mean and median aren't too far apart so the distributions look appropriate for further analysis.)***


(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
sex_class <- table(mydata$SEX, mydata$CLASS)

addmargins(sex_class)
barplot(sex_class, legend.text = c("F", "I", "M"), beside = TRUE, ylab = "Number")

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (The A3 class seems to have the most individuals of any class. As classes are further out from that class they decrease pretty regularly. I think if the assignment were to do a histogram of the counts of CLASS, it would look surprisingly close to a normal distribution. Additionally, the distributions by gender are fairly equal, the I class decreasing as the class gets higher, which is expected if the CLASS variable is a rough sectioning of age in to five categories and I stands for infant.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(nrow(mydata), 200),]
#df[sample(nrow(df), 3), ]
plot(work[, 2:6])

```

-----

##### Section 2: (5 points) Summarizing the data using graphics.

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
mydata$COLOR <- "black"
mydata$COLOR[mydata$CLASS == "A1"] <- "blue"
mydata$COLOR[mydata$CLASS == "A2"] <- "deeppink"
mydata$COLOR[mydata$CLASS == "A3"] <- "coral"
mydata$COLOR[mydata$CLASS == "A4"] <- "cyan"
mydata$COLOR[mydata$CLASS == "A5"] <- "gold"

plot(mydata$WHOLE, mydata$VOLUME, col = mydata$COLOR)
```

(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK, col = mydata$COLOR)
abline(0, (max(mydata$SHUCK/mydata$WHOLE)), col = "black")
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (In the first graph, the different classes seem equally distributed above and below an imaginary line that shows the average value of the values. The lower the class, the closer to the origin. In the second graph, the higher classes are distributed roughly in strata. A1 is still close to the origin as in the first graph, but the higher classes like A5 are closer to the x-axis. Also, since the slope of the line in the second graph is set to the max value of the two variable's ratio, all the values except the max ratio individual are found below the line.  )***

-----

##### Section 3: (8 points) Getting insights about the data using graphs.

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX == "F"], ylab = "Frequency", xlab = "", main = "F")
hist(mydata$RATIO[mydata$SEX == "I"], ylab = "", xlab = "Histogram", main = "I")
hist(mydata$RATIO[mydata$SEX == "M"], ylab = "", xlab = "", main = "M")
boxplot(mydata$RATIO[mydata$SEX == "F"] )
boxplot(mydata$RATIO[mydata$SEX == "I"], xlab = "Boxplots" )
boxplot(mydata$RATIO[mydata$SEX == "M"] )
qqnorm(mydata$RATIO[mydata$SEX == "F"], ylab = "Quartiles", xlab = "", main = "" )
qqnorm(mydata$RATIO[mydata$SEX == "I"], ylab = "", xlab = "Q-Q plots", main = "")
qqnorm(mydata$RATIO[mydata$SEX == "M"], ylab = "", xlab = "", main = "" )
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Visually, the three Q-Q plots do not seem to vary greatly from where the normal line would be. The histograms likewise look fairly bell curve shaped with only a small hint of potential left-skewedness. The boxplots show farily even distribution around the mean with few outliers. Overall, the three groups of mydata$SEX look appreciably normal. To be more sure, you could do a Shaprio-Wilk test of noramality using the built-in shapiro.test() function in R. For this test, if the p-value is less than 0.05, the distribution is significantly different from normal.)***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
ratio_mild.high <- mydata$RATIO[which(1.5*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.75) < mydata$RATIO & mydata$RATIO < 3*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.75))]
ratio_extreme.high <- mydata$RATIO[which(3*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.75) < mydata$RATIO)]

ratio_mild.low <- mydata$RATIO[which(-1.5*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.25) > mydata$RATIO & mydata$RATIO > -3*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.25))]
ratio_extreme.low <- mydata$RATIO[which(-3*IQR(mydata$RATIO)+quantile(mydata$RATIO, 0.25) > mydata$RATIO)]

str(ratio_extreme.high)
str(ratio_mild.high)
str(ratio_mild.low)
str(ratio_extreme.low)

ratio.outliers <- c(ratio_extreme.high, ratio_mild.high, ratio_extreme.low, ratio_mild.low)

kable(mydata[mydata$RATIO %in% ratio.outliers,])
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (This dataset seems to have few outliers. For containing over a thousand values, having only 15 outliers is relatively few, meaning that most values are contained within the IQR of the distribution.  Also, the majority of the outliers are high outliers, meaning that for the population of abalones the RATIO measurement is slightly weighted towards higher ratio individuals. The class of the extreme high-ratio individuals seems to be weighted towards A1 as well.)***

-----

##### Section 4: (8 points) Getting insights about possible predictors.

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
#boxplot1 <- mydata$VOLUME[mydata$CLASS == "A1"] + mydata$WHOLE[mydata$CLASS == "A1"]


par(mfrow = c(2, 2))
boxplot(mydata$VOLUME ~ mydata$CLASS, ylab = "Counts", main = "Volume by Class")
boxplot(mydata$WHOLE ~ mydata$CLASS, ylab = "", main = "Whole by Class")
plot(mydata$VOLUME ~ mydata$RINGS, ylab = "Counts", main = "Volume by Rings")
plot(mydata$WHOLE ~ mydata$RINGS, main = "Whole by Rings")
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (For this answer, I'm assuming the older the abalone, the more rings and the higher the class. I think that these variables work as predictors of age up to a point. As the number of rings increases, so does the WHOLE measurement, up until about 12 rings where the relationship becomes more sporadic. The VOLUME measurement responds the same way. The class variable also steadily increases for both VOLUME and WHOLE until about A5. So for a majority of the data points, the prediction power is useful.)***

-----

##### Section 5: (12 points) Getting insights regarding different groups in the data.

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

vol.mean <- aggregate(mydata$VOLUME, by = list(mydata$SEX, mydata$CLASS), FUN = mean)
shuck.mean <- aggregate(mydata$SHUCK, by = list(mydata$SEX, mydata$CLASS), FUN = mean)
ratio.mean <- aggregate(mydata$RATIO, by = list(mydata$SEX, mydata$CLASS), FUN = mean)

vol.matrix <- matrix(vol.mean$x, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"), c("A1", "A2", "A3", "A4", "A5")))
shuck.matrix <- matrix(shuck.mean$x, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"), c("A1", "A2", "A3", "A4", "A5")))
ratio.matrix <- matrix(ratio.mean$x, nrow = 3, ncol = 5, dimnames = list(c("F", "I", "M"), c("A1", "A2", "A3", "A4", "A5")))

kable(vol.matrix, format = "markdown", caption = "$Volume")
kable(shuck.matrix, format = "markdown", caption = "$Shuck")
kable(ratio.matrix, format = "markdown", caption = "$Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

interaction.plot(mydata$CLASS, mydata$SEX, mydata$RATIO, fun = mean, type = "l", xlab = "Class", ylab = "Ratio Means", trace.label = "Gender")
interaction.plot(mydata$CLASS, mydata$SEX, mydata$VOLUME, fun = mean, type = "l", xlab = "Class", ylab = "Volume Means", trace.label = "Gender")
interaction.plot(mydata$CLASS, mydata$SEX, mydata$SHUCK, fun = mean, type = "l", xlab = "Class", ylab = "Shuck Means", trace.label = "Gender")

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (The first question that springs to mind when looking at the 2nd and 3rd graphs is why are the females (F) both larger in volume and shuck weight than the males? It makes sense that the infant (I) class is smaller but the sexual dimorphism makes me wonder about the environment in which the abalones live. I am also surprised that despite the M / F difference in both volume and shuck weight, the ratio means remain very close for the two groups. I wonder what underlying factor might allow for the differences while maintaing similar means. Maybe something involving nutrient availability. )***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2,2))
boxplot(mydata$VOLUME[which(mydata$SEX == "I")] ~ mydata$RINGS[which(mydata$SEX == "I")], subset = mydata$RINGS < 16, ylim = c(0, 1100), ylab = "Volume", xlab = "Rings", main = "I Volume")

boxplot(mydata$VOLUME[which(mydata$SEX == "M" | mydata$SEX == "F")] ~ mydata$RINGS[which(mydata$SEX == "M" | mydata$SEX == "F")], subset = mydata$RINGS < 16, xlim = c(0, 13), ylim = c(0, 1100), ylab = "Volume", xlab = "Rings", main = "M and F Volume")


boxplot(mydata$WHOLE[which(mydata$SEX == "I")] ~ mydata$RINGS[which(mydata$SEX == "I")], subset = mydata$RINGS < 16, ylim = c(0, 400), ylab = "Whole", xlab = "Rings", main = "I Whole")

boxplot(mydata$WHOLE[which(mydata$SEX == "M" | mydata$SEX == "F")] ~ mydata$RINGS[which(mydata$SEX == "M" | mydata$SEX == "F")], subset = mydata$RINGS < 16, xlim = c(0, 13), ylim = c(0, 400), ylab = "Whole", xlab = "Rings", main = "M and F Whole")


```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (First, regardless of status, the ring growth by either volume or whole looks fairly normally distributed, up to a point. For the infants, the peak seems to be at 10 rings while for adults it seems to be 11. Both groups decrease after this point with more variability as the number of rings increase. The adults also look to have higher means and upper quartiles than the infants.)***

-----

##### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).

**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (Looking at a plot of RINGS vs VOLUME or RINGS VS WHOLE shows the picture of why it failed fairly well, I think. Close to the origin, the younger and smaller abalones are distributed narrowly within a limited range of sizes and weights. Once you go past this initial limit of, say, five rings, the variability explodes in all directions. From this position to the limit of the graph, it looks like the relationship between the measured physical characteristics and number of rings does not follow a strictly linear pattern. Overall, if you boiled the relationship down to a line it would probably be roughly exponential. But if you included boundaries that showed where fifty percent of the data is within, you would still have lots of observations outside of that range. That factor makes it more likely that abalone age does not correspond to the measured characteristics but to outside phenomena that was suggested in the background information (nutrient availability, temperature, etc.). You could generate an r^2 value to see how well the line fits the data, but I would guess that the observed characteristics would be a poor predictor of abalone age at best.)***


**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would first ask what the number of observations are in the study data. Even if the summary and histogram look beautiful, if there aren't a sufficient number of observations you won't be able to derive useful inferences from the study. On the same line of thinking, you could do a power calculation to determine the number of observations needed for the given effect size to see if the study can produce significant findings. I would also ask how the observations were recorded, if there was any bias in the sampling method and similar lines of questioning. If the sampling was biased, the chances of the data being representative of the intended effect are reduced.)***


**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (If the analyst only gets access to the observational data without knowing the sampling strategy or types of analyses desired, it may produce a lopsided view of the data. With greater knowledge of the limitations of the study, the better the analyses might be tailored to show the intended relationships. For example, if a population is known to be right-skewed and your data is more normally distibuted, you might reasonably assume that a certain portion of the population was missed in the sampling. With enough background knowledge you may be able to compensate for irregularities like these. To determine causality from an observational study would be an incredibly hard thing to do. The study would have to be very well constructed to show that the relationship is cotemporal and that the initiating factor is necessary and sufficient for the response. Even under controlled laboratory conditions showing that something causes something is a difficult task. Doing so using only data from an observational study is a further degree more difficult. Even though such a study might not show causation, you might still be able to see and measure the size and duration of the effect you are looking at, which can be useful data for all future studies in the area.)***
